taxize: taxonomic search and retrieval in R

### 2013-10-01 f1000_rev/taxize_appA.Rnw

% \begin{figure}[!h] 
% \centering 
<<phylogeny, eval=TRUE, dev='pdf', out.width='4in', out.height='4in', cache=TRUE, message=FALSE>>=
# Fetch phylogeny from phylomatic 
phylogeny <- phylomatic_tree(taxa=as.character(allnames$spname), taxnames=TRUE, get = 'POST', informat='newick', method = "phylomatic",
    storedtree = "R20120829", taxaformat = "slashpath", outformat = "newick", clean = "true", parallel=TRUE)
# Format teeth-labels 
phylogeny$tip.label <- capwords(phylogeny$tip.label, onlyfirst=TRUE)
# plot phylogeny 
plot(phylogeny)
@
% \caption{a phylogeny...} 
% \end{figure}

@@

<<plot_map, eval=TRUE, dev='pdf', out.width='.9\\textwidth', cache=TRUE, fig.cap='A map', fig.pos='h', fig.lp='fig:map', fig.height=7, fig.width=10>>=
library(rgbif)
library(ggplot2)

>>>>

<<phylogeny, eval=TRUE, dev='pdf', out.width='4in', out.height='4in', cache=TRUE, message=FALSE, fig.cap='A phylogeny.', fig.pos='!h', fig.lp='fig:phylo'>>=
# Fetch phylogeny from phylomatic 
phylogeny <- phylomatic_tree(taxa=as.character(allnames$spname), taxnames=TRUE, get = 'POST', informat='newick', method = "phylomatic",
    storedtree = "R20120829", taxaformat = "slashpath", outformat = "newick", clean = "true", parallel=TRUE)
# Format teeth-labels 
phylogeny$tip.label <- capwords(phylogeny$tip.label, onlyfirst=TRUE)
# plot phylogeny 
plot(phylogeny)
@

@@

<<plot_map, eval=TRUE, dev='pdf', out.width='.9\\textwidth', cache=TRUE, fig.cap='A map.', fig.pos='h', fig.lp='fig:map', fig.height=7, fig.width=10>>=
library(rgbif)
library(ggplot2)

### 2013-10-01 f1000_rev/taxize_appA.tex

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# The tnrs function accepts a vector of 1 or more}
\hlstd{splist_tnrs} \hlkwb{<-} \hlkwd{tnrs}\hlstd{(}\hlkwc{query} \hlstd{= splist,} \hlkwc{getpost} \hlstd{=} \hlstr{"POST"}\hlstd{,} \hlkwc{source_} \hlstd{=} \hlstr{"iPlant_TNRS"}\hlstd{)}
\hlcom{# Remove some fields}
\hlstd{(splist_tnrs} \hlkwb{<-} \hlstd{splist_tnrs[,} \hlopt{!}\hlkwd{names}\hlstd{(splist_tnrs)} \hlopt{%in%} \hlkwd{c}\hlstd{(}\hlstr{"matchedName"}\hlstd{,} \hlstr{"annotations"}\hlstd{,}
    \hlstr{"uri"}\hlstd{)])}
\end{alltt}
\begin{verbatim}
#            submittedName         acceptedName    sourceId score
# 3       Helanthus annuus    Helianthus annuus iPlant_TNRS  0.98
# 1         Pinos contorta       Pinus contorta iPlant_TNRS  0.96
# 4  Collomia grandiflorra Collomia grandiflora iPlant_TNRS  0.99
# 5       Abies magnificaa      Abies magnifica iPlant_TNRS  0.98
# 10       Rosa california     Rosa californica iPlant_TNRS  0.99
# 9         Datura wrighti      Datura wrightii iPlant_TNRS  0.98
# 7       Mimulus bicolour      Mimulus bicolor iPlant_TNRS  0.98
# 8       Nicotiana glauca     Nicotiana glauca iPlant_TNRS     1
# 6          Maddia sativa         Madia sativa iPlant_TNRS  0.97
# 2    Bartlettia scapposa   Bartlettia scaposa iPlant_TNRS  0.98
\end{verbatim}
\begin{alltt}
\hlcom{# Note the scores. They suggest that there were no perfect matches, but they}
\hlcom{# were all very close, ranging from 0.77 to 0.99 (1 is the highest).  Let's}
\hlcom{# assume the names in the 'acceptedName' column are correct (and they should}
\hlcom{# be).}
\hlcom{# So here's our updated species list}
\hlstd{(splist} \hlkwb{<-} \hlkwd{as.character}\hlstd{(splist_tnrs}\hlopt{$}\hlstd{acceptedName))}
\end{alltt}
\begin{verbatim}
#  [1] "Helianthus annuus"    "Pinus contorta"       "Collomia grandiflora"
#  [4] "Abies magnifica"      "Rosa californica"     "Datura wrightii"     
#  [7] "Mimulus bicolor"      "Nicotiana glauca"     "Madia sativa"        
# [10] "Bartlettia scaposa"
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{tsns} \hlkwb{<-} \hlkwd{get_tsn}\hlstd{(}\hlkwc{searchterm} \hlstd{= splist,} \hlkwc{searchtype} \hlstd{=} \hlstr{"sciname"}\hlstd{,} \hlkwc{verbose} \hlstd{=} \hlnum{FALSE}\hlstd{)}
\hlstd{comnames} \hlkwb{<-} \hlkwd{lapply}\hlstd{(tsns, getcommonnamesfromtsn)}
\hlcom{# Unfortunately, common names are not standardized like species names, so}
\hlcom{# there are multiple common names for each taxon}
\hlkwd{sapply}\hlstd{(comnames, length)}
\end{alltt}
\begin{verbatim}
#  [1] 3 3 3 3 3 3 3 3 3 3
\end{verbatim}
\begin{alltt}
\hlcom{# So let's just take the first common name for each species}
\hlstd{comnames_vec} \hlkwb{<-} \hlkwd{do.call}\hlstd{(c,} \hlkwd{lapply}\hlstd{(comnames,} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{)} \hlkwd{as.character}\hlstd{(x[}\hlnum{1}\hlstd{,} \hlstr{"comname"}\hlstd{])))}
\hlcom{# And we can make a data.frame of our scientific and common names}
\hlstd{(allnames} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}\hlkwc{spname} \hlstd{= splist,} \hlkwc{comname} \hlstd{= comnames_vec))}
\end{alltt}
\begin{verbatim}
#                  spname                       comname
# 1     Helianthus annuus              common sunflower
# 2        Pinus contorta                lodgepole pine
# 3  Collomia grandiflora        largeflowered collomia
# 4       Abies magnifica                    golden fir
# 5      Rosa californica           California wildrose
# 6       Datura wrightii            sacred thorn-apple
# 7       Mimulus bicolor yellow and white monkeyflower
# 8      Nicotiana glauca                  tree tobacco
# 9          Madia sativa                 coast tarweed
# 10   Bartlettia scaposa                Bartlett daisy
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# As we already have Taxonomic Serial Numbers from ITIS, let's just get}
\hlcom{# classifications from ITIS. Note that we could use uBio instead.}
\hlstd{class_list} \hlkwb{<-} \hlkwd{classification}\hlstd{(tsns)}
\hlkwd{sapply}\hlstd{(class_list, nrow)}
\end{alltt}
\begin{verbatim}
#  [1] 12 11 12 11 12 12 12 12 12 12
\end{verbatim}
\begin{alltt}
\hlcom{# And we can attach these names to our allnames data.frame}
\hlkwd{library}\hlstd{(plyr)}
\hlstd{gethiernames} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{) \{}
    \hlstd{temp} \hlkwb{<-} \hlstd{x[,} \hlkwd{c}\hlstd{(}\hlstr{"rankName"}\hlstd{,} \hlstr{"taxonName"}\hlstd{)]}
    \hlstd{values} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}\hlkwd{t}\hlstd{(temp[,} \hlnum{2}\hlstd{]))}
    \hlkwd{names}\hlstd{(values)} \hlkwb{<-} \hlstd{temp[,} \hlnum{1}\hlstd{]}
    \hlkwd{return}\hlstd{(values)}
\hlstd{\}}
\hlstd{class_df} \hlkwb{<-} \hlkwd{ldply}\hlstd{(class_list, gethiernames)}
\hlstd{allnames_df} \hlkwb{<-} \hlkwd{merge}\hlstd{(allnames, class_df,} \hlkwc{by.x} \hlstd{=} \hlstr{"spname"}\hlstd{,} \hlkwc{by.y} \hlstd{=} \hlstr{"Species"}\hlstd{)}
\hlcom{# Now that we have allnames_df, we can start to see some relationships among}
\hlcom{# species simply by their shared taxonomic names}
\hlstd{allnames_df[}\hlnum{1}\hlopt{:}\hlnum{2}\hlstd{, ]}
\end{alltt}
\begin{verbatim}
#               spname        comname Kingdom     Subkingdom Infrakingdom
# 1    Abies magnifica     golden fir Plantae Viridaeplantae Streptophyta
# 2 Bartlettia scaposa Bartlett daisy Plantae Viridaeplantae Streptophyta
#       Division     Subdivision Infradivision         Class Superorder
# 1 Tracheophyta Spermatophytina  Gymnospermae     Pinopsida       <NA>
# 2 Tracheophyta Spermatophytina  Angiospermae Magnoliopsida  Asteranae
#       Order     Family      Genus
# 1   Pinales   Pinaceae      Abies
# 2 Asterales Asteraceae Bartlettia
\end{verbatim}
\begin{alltt}

@@

% \begin{figure}[!h] 
% \centering 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# Fetch phylogeny from phylomatic}
\hlstd{phylogeny} \hlkwb{<-} \hlkwd{phylomatic_tree}\hlstd{(}\hlkwc{taxa} \hlstd{=} \hlkwd{as.character}\hlstd{(allnames}\hlopt{$}\hlstd{spname),} \hlkwc{taxnames} \hlstd{=} \hlnum{TRUE}\hlstd{,}
    \hlkwc{get} \hlstd{=} \hlstr{"POST"}\hlstd{,} \hlkwc{informat} \hlstd{=} \hlstr{"newick"}\hlstd{,} \hlkwc{method} \hlstd{=} \hlstr{"phylomatic"}\hlstd{,} \hlkwc{storedtree} \hlstd{=} \hlstr{"R20120829"}\hlstd{,}
    \hlkwc{taxaformat} \hlstd{=} \hlstr{"slashpath"}\hlstd{,} \hlkwc{outformat} \hlstd{=} \hlstr{"newick"}\hlstd{,} \hlkwc{clean} \hlstd{=} \hlstr{"true"}\hlstd{,} \hlkwc{parallel} \hlstd{=} \hlnum{TRUE}\hlstd{)}
\hlcom{# Format teeth-labels}
\hlstd{phylogeny}\hlopt{$}\hlstd{tip.label} \hlkwb{<-} \hlkwd{capwords}\hlstd{(phylogeny}\hlopt{$}\hlstd{tip.label,} \hlkwc{onlyfirst} \hlstd{=} \hlnum{TRUE}\hlstd{)}
\hlcom{# plot phylogeny}
\hlkwd{plot}\hlstd{(phylogeny)}
\end{alltt}
\end{kframe}

@@

@@

% \caption{a phylogeny...} 
% \end{figure}

@@

\caption[A map]{A map\label{fig:mapplot_map}}
\end{figure}

>>>>

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# The tnrs function accepts a vector of 1 or more}
\hlstd{splist_tnrs} \hlkwb{<-} \hlkwd{tnrs}\hlstd{(}\hlkwc{query} \hlstd{= splist,} \hlkwc{getpost} \hlstd{=} \hlstr{"POST"}\hlstd{,} \hlkwc{source_} \hlstd{=} \hlstr{"iPlant_TNRS"}\hlstd{)}
\hlcom{# Remove some fields}
\hlstd{(splist_tnrs} \hlkwb{<-} \hlstd{splist_tnrs[,} \hlopt{!}\hlkwd{names}\hlstd{(splist_tnrs)} \hlopt{%in%} \hlkwd{c}\hlstd{(}\hlstr{"matchedName"}\hlstd{,} \hlstr{"annotations"}\hlstd{,}
    \hlstr{"uri"}\hlstd{)])}
\end{alltt}
\begin{verbatim}
#           submittedName         acceptedName    sourceId score
# 5      Helanthus annuus    Helianthus annuus iPlant_TNRS  0.98
# 1        Pinos contorta       Pinus contorta iPlant_TNRS  0.96
# 7 Collomia grandiflorra Collomia grandiflora iPlant_TNRS  0.99
# 6       Rosa california     Rosa californica iPlant_TNRS  0.99
# 4      Mimulus bicolour      Mimulus bicolor iPlant_TNRS  0.98
# 3      Nicotiana glauca     Nicotiana glauca iPlant_TNRS     1
# 2         Maddia sativa         Madia sativa iPlant_TNRS  0.97
\end{verbatim}
\begin{alltt}
\hlcom{# Note the scores. They suggest that there were no perfect matches, but they}
\hlcom{# were all very close, ranging from 0.77 to 0.99 (1 is the highest).  Let's}
\hlcom{# assume the names in the 'acceptedName' column are correct (and they should}
\hlcom{# be).}
\hlcom{# So here's our updated species list}
\hlstd{(splist} \hlkwb{<-} \hlkwd{as.character}\hlstd{(splist_tnrs}\hlopt{$}\hlstd{acceptedName))}
\end{alltt}
\begin{verbatim}
# [1] "Helianthus annuus"    "Pinus contorta"       "Collomia grandiflora"
# [4] "Rosa californica"     "Mimulus bicolor"      "Nicotiana glauca"    
# [7] "Madia sativa"
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{tsns} \hlkwb{<-} \hlkwd{get_tsn}\hlstd{(}\hlkwc{searchterm} \hlstd{= splist,} \hlkwc{searchtype} \hlstd{=} \hlstr{"sciname"}\hlstd{,} \hlkwc{verbose} \hlstd{=} \hlnum{FALSE}\hlstd{)}
\hlstd{comnames} \hlkwb{<-} \hlkwd{lapply}\hlstd{(tsns, getcommonnamesfromtsn)}
\hlcom{# Unfortunately, common names are not standardized like species names, so}
\hlcom{# there are multiple common names for each taxon}
\hlkwd{sapply}\hlstd{(comnames, length)}
\end{alltt}
\begin{verbatim}
# [1] 3 3 3 3 3 3 3
\end{verbatim}
\begin{alltt}
\hlcom{# So let's just take the first common name for each species}
\hlstd{comnames_vec} \hlkwb{<-} \hlkwd{do.call}\hlstd{(c,} \hlkwd{lapply}\hlstd{(comnames,} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{)} \hlkwd{as.character}\hlstd{(x[}\hlnum{1}\hlstd{,} \hlstr{"comname"}\hlstd{])))}
\hlcom{# And we can make a data.frame of our scientific and common names}
\hlstd{(allnames} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}\hlkwc{spname} \hlstd{= splist,} \hlkwc{comname} \hlstd{= comnames_vec))}
\end{alltt}
\begin{verbatim}
#                 spname                       comname
# 1    Helianthus annuus              common sunflower
# 2       Pinus contorta                lodgepole pine
# 3 Collomia grandiflora        largeflowered collomia
# 4     Rosa californica           California wildrose
# 5      Mimulus bicolor yellow and white monkeyflower
# 6     Nicotiana glauca                  tree tobacco
# 7         Madia sativa                 coast tarweed
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# As we already have Taxonomic Serial Numbers from ITIS, let's just get}
\hlcom{# classifications from ITIS. Note that we could use uBio instead.}
\hlstd{class_list} \hlkwb{<-} \hlkwd{classification}\hlstd{(tsns)}
\hlkwd{sapply}\hlstd{(class_list, nrow)}
\end{alltt}
\begin{verbatim}
# [1] 12 11 12 12 12 12 12
\end{verbatim}
\begin{alltt}
\hlcom{# And we can attach these names to our allnames data.frame}
\hlkwd{library}\hlstd{(plyr)}
\hlstd{gethiernames} \hlkwb{<-} \hlkwa{function}\hlstd{(}\hlkwc{x}\hlstd{) \{}
    \hlstd{temp} \hlkwb{<-} \hlstd{x[,} \hlkwd{c}\hlstd{(}\hlstr{"rankName"}\hlstd{,} \hlstr{"taxonName"}\hlstd{)]}
    \hlstd{values} \hlkwb{<-} \hlkwd{data.frame}\hlstd{(}\hlkwd{t}\hlstd{(temp[,} \hlnum{2}\hlstd{]))}
    \hlkwd{names}\hlstd{(values)} \hlkwb{<-} \hlstd{temp[,} \hlnum{1}\hlstd{]}
    \hlkwd{return}\hlstd{(values)}
\hlstd{\}}
\hlstd{class_df} \hlkwb{<-} \hlkwd{ldply}\hlstd{(class_list, gethiernames)}
\hlstd{allnames_df} \hlkwb{<-} \hlkwd{merge}\hlstd{(allnames, class_df,} \hlkwc{by.x} \hlstd{=} \hlstr{"spname"}\hlstd{,} \hlkwc{by.y} \hlstd{=} \hlstr{"Species"}\hlstd{)}
\hlcom{# Now that we have allnames_df, we can start to see some relationships among}
\hlcom{# species simply by their shared taxonomic names}
\hlstd{allnames_df[}\hlnum{1}\hlopt{:}\hlnum{2}\hlstd{, ]}
\end{alltt}
\begin{verbatim}
#                 spname                comname Kingdom     Subkingdom
# 1 Collomia grandiflora largeflowered collomia Plantae Viridaeplantae
# 2    Helianthus annuus       common sunflower Plantae Viridaeplantae
#   Infrakingdom     Division     Subdivision Infradivision         Class
# 1 Streptophyta Tracheophyta Spermatophytina  Angiospermae Magnoliopsida
# 2 Streptophyta Tracheophyta Spermatophytina  Angiospermae Magnoliopsida
#   Superorder     Order        Family      Genus
# 1  Asteranae  Ericales Polemoniaceae   Collomia
# 2  Asteranae Asterales    Asteraceae Helianthus
\end{verbatim}
\begin{alltt}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# Fetch phylogeny from phylomatic}
\hlstd{phylogeny} \hlkwb{<-} \hlkwd{phylomatic_tree}\hlstd{(}\hlkwc{taxa} \hlstd{=} \hlkwd{as.character}\hlstd{(allnames}\hlopt{$}\hlstd{spname),} \hlkwc{taxnames} \hlstd{=} \hlnum{TRUE}\hlstd{,}
    \hlkwc{get} \hlstd{=} \hlstr{"POST"}\hlstd{,} \hlkwc{informat} \hlstd{=} \hlstr{"newick"}\hlstd{,} \hlkwc{method} \hlstd{=} \hlstr{"phylomatic"}\hlstd{,} \hlkwc{storedtree} \hlstd{=} \hlstr{"R20120829"}\hlstd{,}
    \hlkwc{taxaformat} \hlstd{=} \hlstr{"slashpath"}\hlstd{,} \hlkwc{outformat} \hlstd{=} \hlstr{"newick"}\hlstd{,} \hlkwc{clean} \hlstd{=} \hlstr{"true"}\hlstd{,} \hlkwc{parallel} \hlstd{=} \hlnum{TRUE}\hlstd{)}
\hlcom{# Format teeth-labels}
\hlstd{phylogeny}\hlopt{$}\hlstd{tip.label} \hlkwb{<-} \hlkwd{capwords}\hlstd{(phylogeny}\hlopt{$}\hlstd{tip.label,} \hlkwc{onlyfirst} \hlstd{=} \hlnum{TRUE}\hlstd{)}
\hlcom{# plot phylogeny}
\hlkwd{plot}\hlstd{(phylogeny)}
\end{alltt}
\end{kframe}\begin{figure}[!h]

@@

\caption[A phylogeny]{A phylogeny.\label{fig:phylophylogeny}}
\end{figure}

@@

@@

\caption[A map]{A map.\label{fig:mapplot_map}}
\end{figure}

### 2013-10-01 f1000_rev/taxize_ms.Rnw

A user can use the mentioned \emph{classification}-function to retrieve the taxonomic hierarchy and then search the hierarchies up- and downwards for matches. Here is an example to match a species with names on three different taxonomic levels.

@@

B1 <- "gammarus roeseli"
B2 <- "gammarus"
B3 <- "gammaridae"

A_clas <- classification(A, db = 'ncbi')
B1_clas <- classification(B1, db = 'ncbi')
B2_clas <- classification(B2, db = 'ncbi')
B3_clas <- classification(B3, db = 'ncbi')

B1[match(A, B1)]
A_clas[[1]]$Rank[tolower(A_clas[[1]]$ScientificName) %in% B2]
A_clas[[1]]$Rank[tolower(A_clas[[1]]$ScientificName) %in% B3]
@

If one finds a direct match (here \emph{Gammarus roeseli}), they will be lucky. But you can also match Gammaridae with \emph{Gammarus roeseli}, but on a lower taxonomic level. A more comprehensive and realistic example (matching a trait table with an abundance table) is given in Appendix B.

>>>>

A user can use the mentioned \emph{classification}-function to retrieve the taxonomic hierarchy and then search the hierarchies up- and downwards for matches. Here is an example to match a species (A) with names of on different taxonomic levels (B1 \& B2).

@@

B1 <- "gammarus"
B2 <- "gammaridae"

A_clas <- classification(A, db = 'ncbi')
B1_clas <- classification(B1, db = 'ncbi')
B2_clas <- classification(B2, db = 'ncbi')

A_clas[[1]]$Rank[tolower(A_clas[[1]]$ScientificName) %in% B1]
A_clas[[1]]$Rank[tolower(A_clas[[1]]$ScientificName) %in% B2]
@

If one finds a direct match (here \emph{Gammarus roeseli}), they will be lucky. However, Gammaridae can also be matched with \emph{Gammarus roeseli}, but on a lower taxonomic level. A more comprehensive and realistic example (matching a trait table with an abundance table) is given in Appendix B.
